---
title: "Cultural distance and value mediators with mlcvi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cultural distance and value mediators with mlcvi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mlcvi)
```

The mlcvi package implements the machine learning-based cultural values
inventory (ML-CVI) for cross-cultural research. This vignette walks through
the four things the package does:

1. look up cultural distance between countries (Kogut-Singh, Shulgin, ML-CVI);
2. screen the ML-CVI items as mediators between a binary culture variable and
   an outcome;
3. impute a country-level indicator for countries where it is unobserved,
   using Ridge regression weighted by ML-CVI feature importances;
4. build distance matrices from custom item sets and merge distances into
   panel data.

## Cultural distance between two countries

`mlcvi.get.distance()` takes two ISO alpha-3 codes and a method. The
Kogut-Singh index is computed from Hofstede's dimensions (four- and
six-dimension variants); the Shulgin and ML-CVI methods look up packaged
tables.

```{r}
mlcvi.get.distance("USA", "JPN", method = "KS")
mlcvi.get.distance("USA", "JPN", method = "MLCVI")
```

The function returns a list invisibly, so it can be used programmatically
with `verbose = FALSE`. Problems are reported in an `$error` element rather
than as R errors:

```{r}
res <- mlcvi.get.distance("USA", "XXX", method = "MLCVI", verbose = FALSE)
res$MLCVI$value
res$MLCVI$error
```

## Distances for a panel of country pairs

For dyadic data (trade flows, migration, and so on) it is more convenient to
look up many pairs at once. `mlcvi_get_panel()` appends a `distance` column
and a `distance_note` column to a data frame of pairs, keeping the rows and
their order unchanged so the result merges back one-to-one.

```{r}
pairs <- data.frame(
  iso1 = c("USA", "USA", "DEU", "USA"),
  iso2 = c("JPN", "CHN", "CHN", "XXX"),
  year = c(2005, 2010, 2010, 2015)
)
mlcvi_get_panel(pairs, method = "MLCVI")
mlcvi_get_panel(pairs, method = "KS")
```

The packaged tables are time-invariant, so the same distance applies to every
year (`rule = "static"`). The `rule` argument reserves `"wave"` and `"locf"`
for wave-specific tables in a later version.

## Screening ML-CVI items as mediators

`mlcvi.get.mediator()` runs a Sobel test for each of the 60 ML-CVI items as a
mediator between a binary culture variable and a continuous outcome, adjusts
the p-values, and reports the significant items. The bundled `MLCVI_4A` data
compare United States and Indian respondents on environmental behaviour.

```{r}
out <- mlcvi.get.mediator(df = MLCVI_4A, iv = "US0IN1", dv = "envavg")
out$significant_items
head(out$results[, c("mediator", "indirect", "sobel_z", "p_adj")])
```

Mediator columns are detected by name (`mlcvi1` to `mlcvi60`, or `abc1` to
`abc60`), or can be given explicitly through `mediator_names`.

## Imputing a country-level indicator

`mlcvi_extend()` takes a data frame of country codes and an indicator that is
observed for some countries and `NA` for others. It aggregates the
individual-level WVS training responses to country means, fits a Ridge
regression whose per-item penalties reflect the ML-CVI feature importances,
and predicts the indicator for the missing countries.

The full training matrix is large and is downloaded once on first use. For
this vignette we use the bundled subsample and a fixed penalty so that the
example runs in seconds.

```{r}
codes  <- sub("^s003_", "", colnames(train_output_matrix))
set.seed(1)
scores <- data.frame(s003 = codes, values = rnorm(length(codes)))
scores$values[1:3] <- NA

small <- mlcvi:::mlcvi_train_input(small = TRUE)
filled <- mlcvi_extend(scores, train_input_matrix = small, lambda = 0.1,
                       repeats = 1L, verbose = FALSE)
filled[filled$s003 %in% codes[1:3], ]
```

With the defaults (`lambda = NULL`), the penalty is tuned by Bayesian
optimisation over repeated cross-validation; `mlcvi_ridge_model()` returns
the fitted caret model itself.

## Distance matrices from custom item sets

`mlcvi_build_matrix()` computes classical distances between countries from a
user-specified set of WVS items: a Kogut-Singh style index with items as
dimensions, Euclidean distance, or Mahalanobis distance on the country means.
The result has ISO alpha-3 dimnames and can be passed as `data` to the lookup
functions with `method = "matrix"`, which looks the table up as-is without
attaching a method label.

```{r}
items <- mlcvi_items_default[1:10]
m <- mlcvi_build_matrix(items, data = small, min_n = 20)
m[c("USA", "JPN", "DEU"), c("USA", "JPN", "DEU")]
mlcvi_get_panel(pairs, method = "matrix", data = m)
```

Setting `by_wave = TRUE` returns one matrix per WVS wave. Countries with
fewer than `min_n` respondents in a wave are dropped, and when an item is
unobserved for a country the pairwise distance is renormalized over the items
available to both countries (the `items_used` attribute records how many).

```{r, message = FALSE}
by_wave <- mlcvi_build_matrix(items, data = small, by_wave = TRUE, min_n = 10)
sapply(by_wave, nrow)
```

## Citation

```{r}
citation("mlcvi")
```
